---
title: "Data import with ReadR"
output: html_notebook
---

```{r}
library(tidyverse)
library(rio)
```
Getting started
Most of readr functions are concerned with turning flat files to dataframe

- read_csv() = reads comma delimited files
- read_csv2() = reads semicolon delimited files
- read_tsv() = reads tab delimited files
- read_delim() = reads files with any delim
- read_fwf() = fixed width files
- read_log() = Apache style log files


```{r}
heights = read_csv('heights.csv')
heights2 = import('heights.csv')
heights2
heights
as.tibble(heights2)

#can also supply inline csv
read_csv('a,b,c
         1,2,3
         4,5,6')

#skip some lines
read_csv('The first line of metadata
         The second line
         x,y,z
         1,2,3', skip = 2)

read_csv('#a comment I want to skip
         x,y,z
         1,2,3', comment = '#')

#data might not have column names, you can use col_names = F
read_csv('1,2,3\n4,5,6', col_names = F)

#pass col names
read_csv('1,2,3
         4,5,6', col_names = c('a','b','c'))

#another option is to tweak na which specifies value that are used for missing values

read.csv('a,b,c
         ,1,2,.')
```

Exercise

```{r}
#what function would you use to read a file where fields are sep with |
?read_csv

#aside from file, skip, comment, what other arguments do read_csv and read_tsv have in common

#most important arguments in read_fwf()
?read_fwf

#4 what argument do you need to specify to put the following text to dataframe
read_csv("x,y\n1,'a,b'",quote = "'")

#identify what is wrong in each of the following inline csv
read_csv("a,b1\n1,2,3\n4,4,5,6")
read_csv("a,b,c\n1,2\n1,2,3,4")
read_csv("a,b\n,2\na,b")
```

Parsing a vector - these functions take a character vector and return a more specailized vector
```{r}
str(parse_logical(c('T','F','NA')))
str(parse_integer(c('1','2','3')))
str(parse_date(c('2010-01-01','1979-10-04')))

parse_integer(c('1','231','.','456'), na = '.')

#if parsing fails youll get a warning
x = parse_integer(c('123','345','abc','123.45'))
x
#use problem() to get the complete set
problems(x)
```

The following sections describe parsers in more detail
Numbers
```{r}
parse_double('1.23')
#because readr's default locale is US-centric
parse_double('1,23', local = locale(decimal_mark = ','))

#parse number addresses numeric characters before and after a number. works for currencies and percentages
x = '$100'
y = c('$100','20%','!30!')
typeof(x)
parse_number('$100')
parse_number(x)
parse_number(y)
parse_number('It cost $123.45')

#combination of parse number and locale
parse_number('$123,456,789')
parse_number('$123.456.789', locale = locale(grouping_mark = '.'))
parse_number("123'456'789%%", locale = locale(grouping_mark = "'"))
```
Strings
```{r}
#understand the underlying representation of strings
charToRaw('Hadley')

x1 = 'El Ni\xf1o was particularly bad this year'
x2 = '\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd'

#specify the encoding parse_character
parse_character(x1, locale = locale(encoding = 'latin1'))
parse_character(x2, locale = locale(encoding = 'Shift-JIS'))

#readr has guess encoding to help figure out the encoding
guess_encoding(charToRaw(x1))
guess_encoding(charToRaw(x2))

```
Factors - R uses factors to represent categorical variables that have a known set of possible values.
```{r}
fruit = c('apple','banana')
parse_factor(c('apple','banana','bananana'), levels = fruit)
```

Dates, datetimes, and times

YOu pick between three parsers
- Number of days since 1970
- date time or number of seconds after 1970
- time the number of seconds since midnight
```{r}
#parse_datetime expects ISO8601 date-time. 
parse_datetime('2010-01-01T2010')

#if time is omitted it will be set to midnight
parse_datetime('20101010')

#parse_date expects four digit year, month, date
parse_date('2010-10-01')

#parse time expects hour, minutes, and seconds
library(hms)
parse_time('1:10 am')
parse_time('20:10:01')
```
Datetime format
Year
- %Y 4 digits
- %y 2 digits
Month
- %m 2 digits
- %b abbreviated('Jan')
- %B fullname ('January')
Day
- %d 2 digits
- %e optional leading space
Time
- %H (0-23 hr format)
- %I (0-12, must be used with %p)
- %p (am-pm indicator)
- %M (minutes)
- %S (integer seconds)
- %OS (real seconds)
Non digits
- %. skip one digit non characte
- %* skip any number non digit

```{r}
parse_date('01/01/15', '%m/%d/%y')
parse_date('01/02/15', '%d/%m/%y')
parse_date('01/02/15', '%y/%m/%d')

#if you are using %b or %B with non english names, youll need to set lang argument to locale

parse_date('1 janvier 2015', '%d %B %Y', locale = locale('fr'))
```
Exercises
```{r}
#what are the most important arguments to locale
?locale

#2. What happens if you try to set decimal mark and grouping mark to the same character.what happens to default value of grouping mark when you set the decimal to '.'? what happens to default value decimal mark when u set the grouping mark to '.'
parse_number("$135.435,45", locale = locale(grouping_mark = ".", decimal_mark = ","))

#3. what does dateformat and time_format in locale do
?locale
local_custom = locale(date_format = 'Day %d Mon %m Year %y')
parse_date('Day 01 Mon 02 Year 03', locale = local_custom)

#create a local object that encapsulates settings for types outside US
au_locale = locale(date_format = '%d/%m/%Y')
parse_date('02/01/2006', locale = au_locale)

#parse the ff
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"

parse_date(d1, '%B %d, %Y')
parse_date(d2, '%Y-%b-%d')
parse_date(d3, '%d-%b-%Y')
parse_date(d4, '%B %d (%Y)')
parse_date(d5, '%m/%d/%y')
parse_time(t1, '%H%M')
parse_time(t2, '%I:%M:%OS %p')
```
Parsing a file
readr tries to guess the type of each column
```{r}
guess_parser('2010-01-01')
guess_parser(c('T','F','T','F','T'))

#the defaults however do not work with large files
challenge = read_csv(readr_example('challenge.csv'))

problems(challenge)

#a good strategy is to work column by colum until there is no problem remaining

challenge = read_csv(readr_example('challenge.csv'), col_types = cols(x = col_double(), y = col_date()))

chal = as.tibble(import(readr_example('challenge.csv')))
chal$x = as.double(chal$x)
chal$y = as.Date(chal$y)
tail(chal)
```


Every parse_xyz() function has corresponding col_xyz(). You parse when the data is in character data vector in R already and use col_xyz() when u want to tell readr how to load the data. 

Other strategies for readr
```{r}
#change the maximum row for guessing
challenge2 = read_csv(readr_example('challenge.csv'), guess_max = 1002)
challenge2

#read columns as character vectors
challenge2 = read_csv(readr_example('challenge.csv'), col_types = cols(.default = col_character()))
challenge2
```

writing to a file
```{r}
vignette('rio')

write_csv(challenge, 'challenge.csv')
export(challenge, 'challenge.csv')
```






