---
title: "Vectors"
output: html_notebook
---


```{r}
library(tidyverse)
```

Coercion - implicit coercion happens when you use a logical vector with a numeric summary functions or use a double when integer is expect
```{r}
x = sample(20, 100, replace = T)
y = x > 10
sum(y)
mean(y)

typeof(c(T,1L))
typeof(c(1.5, 1L))

is.atomic(c(T,F,T))
```

Scalars and recycling rules - shorter vector is recycled or repeated 

```{r}
sample(10) + 100

1:10 + 1:2

tibble(x = 1:4, y = 1:2) # error

tibble(x = 1:4, y = rep(1:2, 2))
```

Naming vectors - all vectors can be named
```{r}
a = c(x = 1, y = 2, z = 4)
a['x']

#or use set_names

set_names(1:3, c('a','b','c'))
```

Subsetting - four type of things you can subset vector with
```{r}
#numeric vector containing only integers
x = c('one','two','three', 'four','five')
x[c(1,4,5)]
x[c(1,1,4,4,5,5)]

#negative values drop the element
x[c(-1,-3,-5)]

#subsetting with logical vectors keeps all values correspondign to true
x = c(10,3,NA, 5,8,1, NA)
x[!is.na(x)]

#if you have a named vector you can subset it
x = c('abc'= 1, 'def' = 6, 'xyz'= 11)
x[c('abc','xyz')]

x[] # returns all
```

Exercise
```{r}
#create a function and return the last value

x = c(0,1,3,5,NA, 7, NA)

last_value = function(x){
  x[[length(x)]]
}

last_value(x)

#returns element at even numbered positions

even_elements = function(x){
  x[seq_along(x) %% 2 == 0]
}

even_elements(x)
```

Recursive vectors
```{r}
x = list(1,2,3)
x

#str is a good tool for list because it focused on structure not on contents
str(x)

x_named = list(a = 1, b = 2, c= 3)
x_named
str(x_named)

#list can contain mix of numbers
y = list('a', 1L, 1.5, T)
str(y)

y[[1]]

#list can even contain list

z = list(list(1,2), list(3,4))
str(z)


```

Subsetting with list
```{r}
lis = list(a = 1:3 , b = 'a string', c = pi, d = list(-1,-5))
str(lis)

#[] extracts a sublist
str(a[1:2])
str(a[4])

#[[]] extract components from list
str(y[[1]])
str(lis[[1]])

# $ is a shorthand for extracting named elements in list
lis$b
lis[['b']]


```

Augmented vectors - vectors with additional attributes. 
Factors - 
Date times and times
Tibbles

Factors are designed to represent categorical data that can take fixed set of possible values
```{r}

x = factor(c('ab','cd','ab'), levels = c('ab','cd','ef'))
x
```
tibbles are augmented list. THey have three classes - tbl_df, tbl, and dataframe. They have two attributes (column) names and row.names

```{r}
?tibble
tb = tibble(x = 1:5, y = 5:1)
attributes(tb)

tibble(x = 1:3, y = list("a", 1, list(1:3)))
```






















