---
title: "Relational Data with Dplyr"
output: html_notebook
---


Explore relational data from nycflights13 using verbs from dpylr
```{r}
library(tidyverse)
library(nycflights13)
```
```{r}
nycflights13::airlines
flights
airlines #airline name and code
airports #information about each airports
planes #gives information about the plain, identified by tailnum
weather #weather at each hour
```


Keys - variables used to connect each pair of table. Key is a variable that uniquely identifies an observation. 
Primary key - uniquely identifies an observation on its own table. tailnum in planes is a priamary key
foreign key - identifies an observation in another table. flights tailnum is a foreign key
```{r}
planes %>%
  count(tailnum) %>%
  filter(n  > 1)

weather %>%
  count(year, month, day, hour, origin) %>%
  filter(n > 1)

```

If a table lacks a primary key, it is sometimes useful to add one with mutate and row_number(). This is called surrogate key.

Primary key and corresponding foreign key in another table form a relation.

Exercises
```{r}
## add a surrogate key to flights
flights %>%
  mutate(flight_id = row_number())

#identify the keys of the following datasets
Lahman::Batting %>%
  distinct(playerID) %>%
  count(playerID) %>%
  filter(n > 1)

Lahman::Master
Lahman::Batting
Lahman::Salaries
```

Mutating joins - 
```{r}
f2 = flights %>%
  select(year:day, hour, origin, dest, tailnum, carrier)
f2

#add full airline name to the flights2 data. can combine usign left join

f2 %>%
  select(-origin, - dest) %>%
  left_join(airlines, by = 'carrier')

#you can get same results by using mutate
f2 %>%
  select(-origin, -dest) %>%
  mutate(name = airlines$name[match(carrier, airlines$carrier)])
```

Understanding joins
```{r}
x = tribble(
  ~key, ~val_x,
  1, 'x1',
  2, 'x2',
  3, 'x3'
)

y = tribble(
  ~key, ~val_y,
  1, 'y1',
  2, 'y2',
  4, 'y3'
)

#Inner joins, whenever the keys are equal. Usually not used because its easy to lose observations
x %>%
  inner_join(y, by = 'key')

#Outer joins
#left join keeps observation of x
#right join keeps observation of y
#full join all observation

```

Duplicate keys
```{r}
x = tribble(
  ~key, ~val_x, 
  1, 'x1',
  2, 'x2',
  2, 'x3',
  1, 'x4'
)

y = tribble(
  ~key, ~val_y,
  1, 'y1',
  2, 'y2'
)

x %>%
  left_join(y, by = 'key')

#when you get duplicated keys, you get all possible combinations
x = tribble(
  ~key, ~val_x,
  1, 'x1',
  2, 'x2',
  2 ,'x3',
  3, 'x4'
)
y = tribble(
  ~key, ~val_y,
  1 , 'y1',
  2, 'y2',
  2, 'y3',
  3, 'y4' 
)

left_join(x, y, by = 'key')
```

Defining key columns
```{r}
#the default is by = NULL
#without by argument, it matches all same columns
f2 %>%
  left_join(weather)

#flights and planes have same columns but different meaning, we only want to join by tailnum
f2 %>%
  left_join(planes, by ='tailnum')

#each flight has an origin and destination airport so we specify which one we want to join to
airports
f2 %>%
  left_join(airports, by = c('origin' = 'faa'))
```

Exercises
```{r}
#compute the average delay by destination then join on airports so that you can show the spatial distribution of delay
install.packages('maps')

airports %>%
  semi_join(flights, c('faa' = 'dest')) %>%
  ggplot(aes(x = lon, y = lat))+
  borders('state')+
  geom_point()+
  coord_quickmap()

flights %>%
  select(arr_delay, dest) %>%
  left_join(airports, c('dest' = 'faa')) %>%
  group_by(dest) %>%
  summarise(mean_delay = mean(arr_delay, na.rm = T)) %>%
  inner_join(airports, c('dest'= 'faa'))%>%
  ggplot(aes(lon, lat, color = mean_delay))+
  geom_point()+
  borders('state')+
  coord_quickmap()

#add the location of origin and destianation (lat, lon) to flights
airport_loc = airports %>%
  select(faa, lat, lon)

flights %>%
  select(year:day, hour, origin, dest) %>%
  left_join(airport_loc, by = c('origin' = 'faa')) %>%
  left_join(airport_loc, by = c('dest' = 'faa'), suffix = c('_origin','_dest'))


#is there a relationship between age of plane and delays

flights %>%
  group_by(tailnum) %>%
  summarise(delay = mean(arr_delay, na.rm = T)) %>%
  left_join(planes)

#what weather conditions is there more likely delay

weather
```

Other implementations
inner_join(x,y) = merge(x,y)
left_join(x,y) = merge(x,y, all.x = T)
right_join(x,y) = merge(x,y, all.y = T)
full_join(x,y)= merge(x,y, all.x = T, all.y = T)

Filtering joins - affects the observations not the variables. There are two types
semi_join(x,y) = keeps all observations in x that have a value in y
anti_join(x,y) = drops all observations in x that have a match in y
Semi joins are useful in matching filtered summary tables back to original rows
```{r}
#top10 popular destinations
top_dest = flights %>%
  count(dest, sort = T) %>%
  head(10)

top_dest
#find each flight that went on each of those destination
flights %>%
  filter(dest %in% top_dest$dest)

#it is difficult to extend the approach in multiple variables. for example imagine you found the top10 days with highest average delays. How would you construct filter statement that used year,month,day to match it back to flights

flights %>%
  semi_join(top_dest)

#only existence of a match is important; it does not matter which observation is matched. filtering joins never duplicate rows like mutating joins do

#anti join keeps rows that dont match

flights %>%
  anti_join(planes, 'tailnum') %>%
  count(tailnum, sort = T)

```


Exercise
```{r}
#what does it mean to have a missing tail num. What do tail numbers that dont have matching records in planes have in common

flights %>%
  anti_join(planes, 'tailnum') %>%
  count(carrier)

#flights that have flown at least 1000 flightt

morethan_100 = flights %>%
  group_by(tailnum) %>%
  summarize(count = n()) %>%
  filter(count >= 100)

morethan_100
planes %>%
  semi_join(morethan_100)

install.packages('fueleconomy')
library(fueleconomy)

#combine vehicles and common. Find only the records of the most common models
vehicles %>%
  semi_join(common, by = c('make','model'))

# what does antijoin between flights and airports tell you
anti_join(flights, airports, by = c('dest'= 'faa')) %>%
  count(dest)

#you might expect that theres implicit relationship between plane and airline, because each plane is flown by a single airline.

plane_carrier = flights %>%
  filter(!is.na(tailnum)) %>%
  distinct(tailnum, carrier) %>%
  count(tailnum) %>%
  filter(n > 1)

flights %>%
  semi_join(plane_carrier)

``` 

Set operations
intersect(x,y) = both only both observations of x and y
union(x,y) = return unique observations in x and y
setdiff(x,y) = observations in x but not in y
```{r}
df1 = tribble(
  ~x, ~y,
  1,1,
  2,1
)

df2 = tribble(
  ~x, ~y,
  1,1,
  1,2
)

intersect(df1, df2)
union(df1, df2)
setdiff(df1,df2)
```










