---
title: "Strings with StringR"
output: html_notebook
---


```{r}
library(tidyverse)
library(stringr)
```


```{r}
"this is a string without a"

#to include a literal
'\''

#to see raw contents of string use writeLines()
x = c('\'','\\')
writeLines(x)
```

Base R contains many functions to work with string but well avoid them because they can be inconsistent. Instead well use functions from stringr. 

str_length() gives the number of characters in string
```{r}
str_length(c('a','R for data science',NA))

```
str_ is a prefix for stringr functions

combining strings - use str_c()
```{r}
str_c('x','y')
str_c('x','y', sep = '-')

#missing values are contagious, if u want to print them as NA, use str_replace_na

x = c('abc',NA)
str_c('|-',x,'-|')
str_c('|-',str_replace_na(x),'-|')
str_c('prefix-', c('a','b','c'), '-suffix') #vectorized 

#objects of length 0 are dropped. this is useful in conjuction with f

name = 'hadley'
time_of_day = 'morning'
birthday = T
logic = F

str_c(
  'Good', time_of_day, ' ', name,
  if(birthday) " and HAPPY BIRTHDAY",
  if(logic) 'this is nice', 'what'
)
```

Subsetting strings

Extract parts of string using str_sub() - it takes the start and end arguments that give the inclusive position of substring

```{r}
x = c('apple','banana','pear','orange')
str_sub(x, 1, 5)

#negative numbers will count backwards
str_sub(x, -3,-1)

#you can also use assignment form to modify strings

str_sub(x, 1, 1) = str_to_upper(str_sub(x,1,1))
x
```
Locales - locales affect the rules
```{r}
x = c('apple','banana','eggplant')
str_sort(x)
str_sort(x, locale = 'haw')
```

Exercises
```{r}
?paste

x = c('apple','banana',NA,'cherry','orange')
paste(x)
print(str_c(x))
paste0(x)

#difference between sep and collapse in str_c()
x = c('x','y')
str_c('x','y', collapse = '.')


str_c('x.y', collapse = '.')

#use str_length and str_sub to get the middle character

x ='JANIDEL'
str_sub(x, str_length(x)/2, str_length(x)/2)

#what does str_wrap() do
?str_wrap

x = 'a very nice weather right here'
str_wrap(x)

#what does str_trim do?
?str_trim
x = '               what is this         this is blank'
str_trim(x)
str_trim(x, side = 'left')
str_squish(x)


#write a function that turns a vector into string with and on the last

str_modifier = function(vector){
  new = ''
  for (i in length(vector){
    vector
  }
}

x = c('a','b','c')

str_c(x[seq_len(length(x)-1)], ' ')
```

Matching patterns with regular expression
str_view() and str_view_all() takes a character vector and a regular expression and show you how they match. 
```{r}
#basic matches
x = c('apple','banana','pear')
str_view(x, 'an')

# '.' matches any character except new line
str_view(x, '.a.')



#to create the regular expression
dot = '\\.'
writeLines(dot)

str_view(c('abc','a.c','bef'), 'a\\.c')
```

Anchors
^ - match the start of the string
$ - match the end of the string
```{r}
x = c('apple','banana','pear')
str_view(x, '^a')
str_view(x, 'a$')

#to force a regular expression to only match a complete string, anchor it with both ^ and $
x = c('apple pie','apple','apple cake')
str_view(x, 'apple')
str_view(x, '^apple$')
```
Exercise
```{r}
#How would you match the literal string '$^$'

str_view(c('$^$', 'as$^$ba'), '$\\$\\^\\$$', match = TRUE)

#using words, find all starts in y

str_view(stringr::words, '^y', match = T)

#ends in y
str_view(stringr::words, 'x$', match = TRUE)

#exactly 3 letters long

str_view(stringr::words, '^...$', match = TRUE)

#have seven letters or more
str_view_all(stringr::words, '.......+', match = TRUE)
```

Character classes and alternatives
- \d any digits
- \s whitespace
- [abc] a,b, or c
- [^abc] anything except abc
- 
```{r}
str_view(c('gray','grey'), 'gr(e|a)y')
```

Exercise
```{r}
#starts with a voewl
str_subset(stringr::words, '^[aeiou]')

#words that contain only consonants
#get words that haa vowel, then negate
str_subset(stringr::words, '([aeiou])', negate = T)

#can also use match = False
str_view(stringr::words, '[aeiou]', match = FALSE)

#end with ed but not eed

str_subset(stringr::words, '([^e])ed$')

#ending with ing or ise
str_subset(stringr::words, '(i(ng|se))$')

#match phone numbers in ph

ph = c('09953819895','123','09265844534','02345678902','0123')
str_subset(ph, '(09)(\\d\\d\\d\\d\\d\\d\\d\\d\\d)')


```
Repetition - 
? 0 or 1
+ 1 or more
* or more
```{r}
x = '1888 is the longest roman numerals MDCCCLXXVIII'
str_view(x, 'CC?')
str_view(x, 'CC+')
str_view(x, 'C[LX]+')
```

Number of matches
{n} = exactly n
{n,} n or more
{,m} at most m
{n,m} between n and m

```{r}
str_view(x, 'C{2}')
str_view(x, 'C{2,}')

```

#Exercise
```{r}
#start with 3 consonants

str_subset(stringr::words, '(^[^aeiou]{3})')

#have three or more bowels in a row
str_subset(stringr::words, '([aeiou]{3,})')
```

Grouping and back references - 
```{r}
#find fruits that have repeated pair of letters
str_subset(fruit, '(..)\\1')
```

(.)\1\1 the same character repeating 3x - aaa
(.)(.)\\2\\1 a pair of characters followed by same pair of characters in reversed order - abba
(..)\1 any two characters repeated
(.).\\1.\\1 a character followed by any character, the original character, any character, and original cahracter again. 
(.)(.)(.).*\\3\\2\\1 3 characters followed by zero or more characters followed by same three characters


```{r}
str_subset(stringr::fruit, '(..)\\1')

#start and end with the same character

str_subset(words, '^(.).*\\1$')

#contain repeated pair of letters
str_subset(words,'(..).*\\1')

#letter repeated in atleast 3 places
str_subset(words, '([a-z]).*\\1.*\\1')
```

Detect matches - returns logical
```{r}
x = c('apple','banana','pear')
str_detect(x, 'e')

#how many words start with t
sum(str_detect(words, '^t'))

#what proportion of words starts with vowel
mean(str_detect(words, '^[aeiou]'))

#find words that dont contain vowesl
#find all words that contain vowels and negate
no_vowels = str_detect(words, '[aeiou]', negate = T)

no_vowels1 = !str_detect(words, '[aeiou]')

words[no_vowels1]

#find the words consisting of only consonants
no_vowels2 = str_detect(words, '^[^aeiou]+$')
words[no_vowels2]

#common use of str_detect() is to select elements that match a pattern
words[str_detect(words,'x$')]

#typically, your strings will be one column of data so you will use filter
df = tibble(
  word = words,
  i = seq_along(words)
)
df
?seq_along

df %>%
  filter(str_detect(word, 'x$'))

#a variation of str_detect is str_count which counts how many matches the are
x = c('apple','banana','pear')
str_count(x, 'a')

#its natural to use str_count with mutate
df %>%
  mutate(vowels = str_count(word, '[aeiou]'),
         consonants = str_count(word, '[^aeiou]'))
```

Exercise
```{r}
#find all words that start or end with x
words[str_detect(words, '^x|x$')]

#start with vowel and ends with constant
words[str_detect(words,'^[aeiou]')&str_detect(words,'[^aeiou]$')]

#atleast one of each different vowels
words[str_detect(words,'a') &
        str_detect(words, 'e') &
        str_detect(words, 'i')]

#what word has the highest number of bowel
vowels = str_count(words, '[aeiou]')
?which
words[which(vowels == max(vowels))]
words[vowels == max(vowels)]
```

Extract matches - use str_extract
```{r}
sentences

#sentences that contains color
colors = c('\\sred\\s','orange','blue','green','purple')
color_match = str_c(colors, collapse = '|')

has_color = str_subset(sentences, color_match)
has_color

#extract what color
str_extract(has_color, color_match)

#note that str_extract matches only the first match

more = sentences[str_count(sentences, color_match) > 1]
str_view(more, color_match)

str_extract(more, color_match)

x = c('a','a b','a b c')
str_extract_all(x, '[a-z]', simplify = T) #returns a matrix


color_match2 <- str_c("\\b(", str_c(colors, collapse = "|"), ")\\b")
color_match2
```
Exercise
```{r}
#extract first word of each sentece

head(sentences)
str_extract(sentences, '[A-z]+')

#find all words that ends with ing
str_extract(sentences, '[A-z]+ing')
```
Grouped matches - 
```{r}
noun = '(a|the) ([^ ]+)'
has_noun = sentences %>%
  str_subset(noun)

#str match gives individual components
has_noun %>%
  str_match(noun)
```


Exercise
```{r}
#Find all words that come after a number

num = '(\\sone\\s|two|three|four|five|six|seven|eight|nine) ([^ ]+)'
has_num = sentences %>%
  str_subset(num)

has_num %>%
  str_match(num)

#find all contractions and seperate pieces

pattern = "([A-Za-z]+)\\'([a-z]+)"
has_apos = sentences %>%
  str_subset(pattern)

has_apos %>%
  str_match(pattern)
```

replacing matches - str_replace and str_replace_all allows to replace matches with new strings
```{r}
x = c('apple','banana','pear')
str_replace_all(x, '[aeiou]', '-')

#flip the order of sentences
str_replace(sentences, '([^ ]+) ([^ ]+) ([^ ]+)', '\\3 \\2 \\1')
```
Exercise
```{r}
#replace all "'" with backlashes

str_replace_all(sentences, "\'", "\\\\")

#implement a version of str_to_lower() using replace_all
replacements <- c("A" = "a", "B" = "b", "C" = "c", "D" = "d", "E" = "e",
                  "F" = "f", "G" = "g", "H" = "h", "I" = "i", "J" = "j", 
                  "K" = "k", "L" = "l", "M" = "m", "N" = "n", "O" = "o", 
                  "P" = "p", "Q" = "q", "R" = "r", "S" = "s", "T" = "t", 
                  "U" = "u", "V" = "v", "W" = "w", "X" = "x", "Y" = "y", 
                  "Z" = "z")
lower_words <- str_replace_all(words, pattern = replacements)
head(lower_words)

#switch the first and last letter in words
str_replace(words,'([a-z])(.*)([a-z])', '\\3\\2\\1')

```


Splitting - split string into pieces
```{r}
sentences %>%
  head(5) %>%
  str_split(' ', simplify = T)

#request a maximum number of pieces
fields = c('Name:Hadley', 'Country: NZ', 'Age:35')
fields %>%
  str_split(':', simplify = T) %>%
  as.data.frame()

#can also split by patterns, character, line, sentence, and word boundary

x = 'This is a sentence. This is another sentence'
str_view_all(x, boundary('sentence'))
str_split(x, boundary('word'))
```

Exercise
```{r}
# split up string like 'apples, pears, and bananas' into individual components

x = 'apples, pears, and bananas'
str_split(x, ', +(and)+')

#what does splitting empty space do, do individual characters
str_split("ab. cd|agt", "")
```
Find matches - str_locate() and str_locate all gives you the starting and ending position of each match.
```{r}

#regular call
str_view(fruit, 'nana')

#shorthand for
str_view(fruit, regex('nana'))

#regex has other arguments
bananas = c('banana','Banana','BANANA','BanAnA')
str_view(bananas, 'nana')
str_view(bananas, regex('nana', ignore_case = T))

#multiline argument allows ^ and $ to match the start and end of each line rather than start and end of complete string

x = 'Line 1\nLine 2\nLine 3'

str_extract_all(x, '^Line')[[1]]

str_extract_all(x, regex('^Line', multiline = T))

#comments = T allows the use of whitespace to make complex regular expressions.
phone = regex('
              \\(? # optional opening parenthesis
              (\\d{3}) # area code
              [)- ]? # optional closing parenthesis
              (\\d{3}) # another three numbers
              [ -]? # optional space or dash
              (\\d{3}) #three more numbers',
              comments = T)

str_match('514-791-8141', phone)
```

Other functions
fixed() - matches specified sequence of bytes. 
coll() - compares strings using standard collation rules

Exercise
```{r}
#most common words in sentences

tibble(word = unlist(str_extract_all(sentences, boundary('word')))) %>%
  mutate(word = str_to_lower(word)) %>%
  count(word) %>%
  arrange(desc(n)) %>%
  mutate(rank = rank(desc(n))) %>%
  filter(rank < 6)
```

Other uses of regular expressions
```{r}
#appropos - searches all objects available in global environment
apropos('replace')

#dir list all files in director
dir(pattern = '.*\\.Rmd$')
```










