---
title: "Exploratory data analysis"
output: html_notebook
---


```{r}
library(tidyverse)
?diamonds

```

Variation = tendecy of values of a variable to change from measurement to measurement. Categorical variables can vary   if you measure across different subject. Every variable has its own pattern of variation, which reveal interesting information. Best way to understand that pattern is to vizualize the distribution

Visualizing distribution
```{r}
# to examine distribution of categorical variable use bar chart
ggplot(data = diamonds)+
  geom_bar(aes(x = cut))

#you can compute the height of each cut manually using dplyr::count()
diamonds %>%
  count(cut)
table(diamonds$cut)

#examine a distribution of a continous variable using histogram
ggplot(data = diamonds)+
  geom_histogram(aes(x = carat, binwidth = .5))

##can compute this by combining dplyr::count() and ggplot2::cut_width()
diamonds %>%
  count( count = cut_width(carat, .5))

#check diamonds less than 3 cards
smaller = filter(diamonds, carat < 3)

ggplot(data = smaller, aes(x = carat))+
  geom_histogram()

#if you wish to overlay multiple histograms in the same plot, use geom_freqpoly(). it performs same calculation as histogram but uses line instead. it is easier to understand overlapping lines than bars

ggplot(data = smaller, mapping = aes(x = carat, color = cut))+
  geom_freqpoly(binwidth = .05)
```

Typical values - 

```{r}
#The following histogram suggest several interesting questions
#Why are there more diamonds at whole carats and common fraction of carst
#why are there more diamonds slightly to the right of each peak than there are slightly to the left of each peak
#why are there no diamonds bigger than 3 carats

ggplot(data = smaller, mapping = aes(x = carat))+
  geom_histogram(binwidth = .01)

#histogram of eruptions in faithful geyser yellowstone
ggplot(data = faithful, mapping = aes(x = eruptions))+
  geom_histogram(binwidth = .25)


```

Unusual values - It is good practice to repeat analysis without outliers. 
```{r}
ggplot(diamonds)+
  geom_histogram(mapping = aes(x = y), binwidth = .5)

#it is hard to see the outliers. We need to zoom in to small values in y axis with coord_cartesian()
ggplot(diamonds)+
  geom_histogram(mapping = aes(x= y), binwidth = .5)+
  coord_cartesian(ylim = c(0,50)) #set ylimit to be 0 to 50

#pick out the unusual values
diamonds %>%
  filter(y < 3 | y>20) %>%
  arrange(y)

#these must be errors because these valeus are impossible

```

Exercises
```{r}
#explore the distribution of x,y, and z variables.
summary(select(diamonds, x,y,z))

#there seems to be outliers because the median value and 3rd quartile is less than 10. Try to remove outliers
filter(diamonds, (x - mean(x)) / sd(x)<3, (x - mean(y)) / sd(y)<3, (z - mean(z)) / sd(z)<3) %>%
  ggplot()+
  geom_freqpoly(aes(x = x), binwidth = .2, color = 'blue')+
  geom_freqpoly(aes(x = y), binwidth = .2, color = 'red')+
  geom_freqpoly(aes(x = z), binwidth = .2, color= 'green')

#most are right skewed

#explore the distribution of price.
summary(diamonds$price)
ggplot(data = diamonds, mapping = aes(x = log(price)))+
  geom_histogram()

#how many diamonds are 0.99 carat and 1 carat, what made the difference
diamonds %>%
  filter(carat == .99 | carat == 1) %>%
  count(carat)


#compare and contrast coord_cartesian() versus xlim, ylim when zooming histogram

ggplot(data = diamonds)+
  geom_histogram(mapping = aes(x = y))+
  ylim(c(0,50))

ggplot(data = diamonds)+
  geom_histogram(mapping = aes(x = y))+
  coord_cartesian(ylim = c(0,50))

#better to use coord cartesian if you just want to zoom
```


Missing values
```{r}
#You have two options
#option 1 : drop the entire row with strange values. Cons is you will lose data
diamonds %>%
  filter(between(y, 3, 20))

# replace unusual values with missing data using mutate and ifelse

diamonds2 = diamonds %>%
  mutate(y = ifelse(y<3 | y>20, NA, y))

#ggplot removes missing values
ggplot(data = diamonds2, mapping = aes(x = x, y = y))+
  geom_point(na.rm = T)

#other times you want to understand what makes observations with missing values.
flights %>%
  mutate(cancelled = is.na(dep_time), 
         sched_hour = sched_dep_time %/% 100,
         sched_min = sched_dep_time %%100, 
         sched_dep_time = sched_hour + sched_min / 60) %>%
  ggplot(mapping = aes(sched_dep_time)) +
  geom_freqpoly(mapping = aes(color = cancelled), binwidth = .25)
```
Exercise
```{r}
#what happens to missing value in barchart, in histogram. Why is there a difference

ggplot(data = diamonds2, aes(x = y))+
  geom_histogram(binwidth = .5)

#missing value is category in barplot
diamonds %>%
  mutate(cut = if_else(runif(n()) < 0.1, NA_character_, as.character(cut))) %>%
  ggplot() +
  geom_bar(mapping = aes(x = cut))
```


Covariation - describes behavior between variables. Covariation is the tendency for two values to vary together. 
```{r}
#categorical  - geom_freqpoly is not useful because the height is given by the count.  Check how price varies with quality

ggplot(data = diamonds, mapping = aes(x = price))+
  geom_freqpoly(mapping = aes(color = cut))

ggplot(data = diamonds, mapping = aes(x = cut))+
  geom_bar()

#to make comparison easier, we need to swap count to density. which is a value standardize so that the area under the curve is 1

ggplot(data = diamonds, mapping = aes(x = price, y = ..density..))+
  geom_freqpoly(mapping = aes(color = cut))

#theres a lot of going in the plot. An alternative to display distribution of a continous variable varied by categorical is boxplot. 

ggplot(data = diamonds, mapping = aes(y = price, x = cut))+
  geom_boxplot()

#the better the quality the cheaper the price on average

#cut is an ordered factor, fair is worse than good. use reorder function
ggplot(data = mpg, mapping = aes(x = class, y = hwy))+
  geom_boxplot()

#reorder class based on median value of hwy
ggplot(data = mpg)+
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))+
  coord_flip()
  
?reorder
```
Exercise
```{r}
#Use what you have learned to improve the visualization of cancelled vs non cancelled flights
flights %>%
  mutate(cancelled = is.na(dep_time), 
         sched_hour = sched_dep_time %/% 100,
         sched_min = sched_dep_time %%100, 
         sched_dep_time = sched_hour + sched_min / 60) %>%
  ggplot(mapping = aes(sched_dep_time, color = cancelled)) +
  geom_density()

#install ggstance package. create a horizontal boxplot. How does this compare to using coordflip
install.packages('ggstance')
library(ggstance)

ggplot(data = mpg)+
  geom_boxploth(mapping = aes( y = reorder(class, hwy, FUN = median), x = hwy))

#boxplots were developed in ERA of small dataset. One approach to remedy this is the letter value plot. install lvplot package and try using geom_lv to display the distribution of price versus cuts

install.packages('lvplot')
library(lvplot)

ggplot(data = diamonds, mapping = aes(y = price, x = cut))+
  geom_lv(aes(fill = ..LV..))

ggplot(data = diamonds, mapping = aes(y = price, x = cut))+
  geom_boxplot()

#compare and contrast geom_violin with faceted geom_histogram or a colored geom_freqpoly
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..))+
  geom_freqpoly(mapping = aes(color = cut))

ggplot(data = diamonds, mapping = aes(x = price))+
  geom_histogram()+
  facet_wrap(~cut, ncol = 1, scales = 'free_y')

ggplot(data = diamonds, mapping = aes(x = cut, y = price))+
  geom_violin()+
  coord_flip()

#if you have a small dataset, sometimes its useful to use geom_jitter() to see relationshipp between continous and categorical. The ggbeeswarm package provides a number of methods similar to geom_jitter.
install.packages('ggbeeswarm')
library(ggbeeswarm)

ggplot(data = mpg)+
  geom_quasirandom(mapping = aes(x = class, y = hwy))

ggplot(data = mpg)+
  geom_beeswarm(mapping = aes(x = class, y = hwy))
```

Two categorical variables - to visualize covariation between categorical variables, youll need to count the number of observations for each combination.
```{r}
ggplot(data = diamonds)+
  geom_count(mapping = aes(x = cut, y = color))
#the size of each circle in the plot displays how many observations occured at each combination of  values. 

#another approach is to use compute count of dplyr
diamonds %>%
  count(color, cut)

#then visualize with geom_tile
diamonds %>%
  count(color, cut) %>%
  ggplot(mapping = aes(x = color, y = cut))+
  geom_tile(mapping = aes(fill = n))

#for larger plots you might want to use heatmaply or d3heatmap
```
Exercise
```{r}
#how would you rescale the count dataset to clearly show the distribution of cut within color or color within cut
table(diamonds$cut, diamonds$color)

diamonds %>%
  count(color, cut) %>%
  mutate(prop = n/ sum(n)) %>%
  ggplot(mapping = aes(x = color, y = cut))+
  geom_tile(mapping = aes(fill = prop))

#use geomtile togethery with dplyr to explroe how average delays vary by destination and month of year.
flights %>%
  group_by(dest, month) %>%
  summarise(delay_mean = mean(dep_delay, na.rm = T),
            n = n()) %>%
  mutate(sum_n = sum(n)) %>%
  select(dest, month, delay_mean, n, sum_n) %>%
  as.data.frame() %>%
  filter(dest == 'ABQ') %>%
  filter(sum_n > 30) %>%
  ggplot(aes(x = as.factor(month), y = dest, fill = delay_mean))+
  geom_tile()

```

Two continous variable
```{r}
#scatterplot

ggplot(data = diamonds)+
  geom_point(mapping = aes(x = carat, y = log(price)), alpha = .2)

#geom_bin2d and geom_hex to bin in two dimensions
ggplot(data = smaller)+
  geom_bin2d(mapping = aes(x = carat, y = price))

#install.packages('hexbin')
library(hexbin)

ggplot(data = smaller)+
  geom_hex(mapping = aes(x = carat, y = price))

#another option is to bin one continous variable so it acts like a categorical variable. Then use boxplot
ggplot(data = smaller, mapping = aes(x = carat, y = price))+
  geom_boxplot(aes(group = cut_width(carat, .2)), varwidth = T)

#make the width of the boxplot proportional to the number of points with varwidth = True

#another approach is to display approximatley the same number of points each bin using cut number

ggplot(data = smaller, aes(x = carat, y = price))+
  geom_boxplot(aes(group = cut_number(carat, 10)))
```
```{r}
table(cut_width(diamonds$price, width = 2500))
```


#Exercise
```{r}
#instead of summarizing conditional distribution with boxplot, you could use a frequency polygon, what do you need to consider when using cut_width versus cut_number. How does that impact visualization of 2D distribution of carat and price


#split carats into5 groups
ggplot(data = diamonds, mapping = aes(x = price))+
  geom_freqpoly(mapping = aes(color = cut_number(carat, 5)))+
  labs(x = 'price', y = 'count', color = 'carat')

#visualize the distribution of carat partitioned by price
ggplot(data = diamonds, aes(x = carat))+
  geom_boxplot(aes(group = cut_number(price, 10)))

#combine two techniques youve learned to visualize combined distribution of cut, price, and carat

ggplot(diamonds, aes(x = carat, y = price))+
  geom_bin2d()+
  facet_wrap(~cut, ncol = 2)

ggplot(diamonds, aes(x = cut_number(carat, 5), y = price))+
  geom_boxplot(aes(color = cut))



```
Patterns and models
```{r}
#scatterplot of old faithful eruptions show two clusters
ggplot(data = faithful)+
  geom_jitter(mapping = aes(x = eruptions, y = waiting))

#models are tools for extracting patterns out of data. Its possible to use a model to remove the very strong relationship between price and carat so we can explore the subtleties that remain. 
#the following code fits a model that predicts price from carat then computes the residuals to give us a view of hte price of diamond once the effect of carat has been removed

library(modelr)

mod = lm(log(price)~log(carat), data = diamonds)
mod

diamonds2 = diamonds %>%
  add_residuals(mod) %>%
  mutate(resid = exp(resid))

ggplot(data = diamonds2)+
  geom_point(mapping = aes(x = carat, y = resid))
#once you removed the strong relationship between carat and price, you can see what you expect in the relationship between cut and price ~ relative to their size, better quality diamonds are more expensive

ggplot(data = diamonds2)+
  geom_boxplot(mapping = aes(x = cut, y = resid))
```




