---
title: "Tidy Data with tidyr"
output: html_notebook
---

```{r}
library(tidyverse)
library(ggplot2)
?count
```

Tidy data
```{r}
table1
table2
table3
table4a
table4b
```


```{r}
#compute rate per 10000
table1 %>%
  mutate(rate = cases/population * 10000)

#compute cases per year
table1 %>%
  group_by(country,year) %>%
  summarize(n = mean(cases))

table1 %>%
  count(year, wt = cases)

#visualize changes overtime
ggplot(table1 , aes(x = year, y = cases))+
  geom_line(aes(group = country), color = 'grey50')+
  geom_point(aes(color = country))
```



Exercises
```{r}
#compute the rate for table2, table4a, table4b
#using table2
cases = table2 %>%
  filter(type == 'cases') %>%
  select(-type) %>%
  rename( 'cases' = 'count')

population = table2 %>%
  filter(type == 'population') %>%
  select(-type) %>%
  rename( 'population' = 'count')

table2_tidy = merge(cases, population) %>%
  mutate(rate = cases/population * 10000)

table1
table2 %>%
  filter(type == 'cases') %>%
  ggplot(mapping = aes(x = year, y = count, group = country))+
  geom_line()+
  geom_point(aes(color = country))
```
Spreading and gathering.
Gathering - some column names are not variables but values. Can use gather
spreading - observation is scattered across multiple row
```{r}
#gathering
table4a

tidy4a = table4a %>%
  gather('1999', '2000', key = 'year', value = 'cases')

tidy4b = table4b %>%
  gather('1999', '2000', key = 'year', value = 'population')

left_join(tidy4a, tidy4b)
merge(tidy4a, tidy4b)


#spreading
table2 %>%
  spread(key = type, value = count)
```

Exercises
```{r}
#why are gather and spread not perfectly symmetrical. both spread and gather have a convert argument, what does it do
stocks = tibble(
  year = c(2015, 2015, 2016,2016),
  half = c(1,2,1,2),
  return = c(1.88, 0.59, 0.92, .17)
)
?spread
stocks %>%
  spread(year, return, convert = T) %>%
  gather('year','return', `2015`:`2016`, convert = T)
#convert for datatype

#why does this code fail
table4a %>%
  gather(`1999`, `2000`, key = 'year',value = 'cases')

#why does spreading this tibble fail? How could you add a new column to fix the problem
people = tribble(
  ~name, ~key, ~value,
  'Philipp Woods', 'age', 45,
  'Philipp Woods', 'height', 186,
  'Philipp Woods','age', 50,
  'Jessica Cordero','age',37,
  'Jessica Cordero','height',156
)
#add a row of distinct observation
people2 = people %>%
  group_by(name, key) %>%
  mutate(obs = row_number())

people2 %>%
  spread(key = key, value = value)

#tidy the simple tibble below
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes", NA, 10,
  "no", 20, 12
)
preg %>%
  gather('male','female', key = 'sex', value = count) %>%
  filter(!is.na(count))
  
```


Separating and  Pull
Separate - pulls one column and split where a character appears
```{r}
table3
table3 %>%
  separate(rate, into = c('cases','population'), sep = '/', convert = T)

#can also pass a vector of integers to sep. will interpret the integers as position of split
table3 %>%
  separate(year, into = c('century','year'), sep = 2)
```

Unite - inverse of separate. Combines multiple columns into single colunn
```{r}
?unite
table5 %>%
  unite(new, century, year, sep = '')

table5 %>%
  unite(new, century, year, sep = '', remove = F)
```

Exercises
```{r}
?separate
#what do fill and extra do in separate
tibble(x = c('a,b,c', 'd,e,f,g','h,i,j')) %>%
  separate(x, into = c('one','two', 'three','four'), fill = 'left')

#combines the two
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"), extra = "merge")

#compare and contrast separate and extract

#separate by character
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
    separate(x, c("variable", "into"), sep = "_")

#by position
tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  separate(x, c("variable", "into"), sep = c(1))

#function extract uses regular expression to specify groups. more flexible than separate because it does not require common separator

tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
   extract(x, c("variable", "id"), regex = "([A-Z])_([0-9])")

tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])([0-9])")


```

Missing values - a value could be missing in two ways
Explicitly = flagged as NA
implicitly = not represented in the data
```{r}
stocks = tibble(
  year = c(2015,2015,2015,2015,2016,2016,2016),
  qtr = c(1,2,3,4,2,3,4),
  return = c(1.88, .59, .35, NA, .92, .17, 2.66)
)
#there are missing
stocks %>%
  spread(year, return)

?gather
stocks %>%
  spread(year, return) %>%
  gather(key = year, value = return, `2015`:`2016`, na.rm = T)

#complete takes a set of columns and finds all unique combinations
stocks %>%
  complete(year, qtr)

treatment = tribble(
  ~person, ~treatment, ~response,
  'Derrick Whitmore', 1, 7,
  NA, 2,10,
  NA, 3,9,
  'Katherine Burke', 1, 4
)
treatment

?fill
treatment %>%
  fill(person, .direction = 'up')
```

Exercises
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

?spread
stocks %>%
  spread(key = year, value = return, fill = 0)
```

Case study
```{r}
summary(who)
who

#gather columns
who1 = who %>%
  gather(new_sp_m014:newrel_f65, key = 'key', value = 'cases', na.rm = T)

#get hint of the structure of values using key
who1 %>%
  count(key)
```


The dataset
- first three letters denote whether column contains new or old cases of tb
- nex two letters describe the type of tb
- sixth letter gives male or female
- remaining numbers are the age group

```{r}
#minor fix on names so that they have consistent names
who2 = who1 %>%
  mutate(key = stringr::str_replace(key, 'newrel','new_rel'))

#separate the codes
who3 = who2 %>%
  separate(key, into = c('new','type','sexage'), sep = '_')

#drop columns
who4 = who3 %>%
  select(-iso2, -iso3, - new)

#split sex and age
who5 = who4 %>%
  separate(sexage, into = c('sex','age'), sep = 1)
who5
```

One line of code
```{r}
tidywho = who %>%
  gather(new_sp_m014:newrel_f65, key = 'code', value = 'value', na.rm = T) %>%
  mutate(code = stringr::str_replace(code, 'newrel','new_rel')) %>%
  separate(code, into = c('new','var','sexage')) %>%
  separate(sexage, into = c('sex','age'), sep = 1) %>%
  select(-iso2, -iso3, -new)

tidywho
```

Exercises
```{r}
#make informative visuzalization of data
tidywho %>%
  group_by(country) %>%
  summarize(mean_cases = mean(value, na.rm = T)) %>%
  arrange(desc(mean_cases)) %>%
  mutate(rank = rank(desc(mean_cases))) %>%
  filter(rank < 11) %>%
  ggplot(aes(country, mean_cases))+
  geom_col()+
  labs(y = 'cases', title = 'top 10 countries with the most number of TB')
```













