---
title: "Functions"
output: html_notebook
---

Writing functions has advantages over using copy paste
1. makes it easier to understand
2. you only need to update code in one place
3. eliminate chance of making incidental mistakes


When should you write a function - whenever you copied or pasted a block of code more than twice
```{r}
df = tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df$a = (df$a - min(df$a, na.rm = T)) / max(df$a, na.rm = T) - min(df$a, na.rm = T)

```


3 key steps to creating a new function

1. Pick a name of function
2. list inputs or arguments to the function inside the function
3. you place the code in the body of the function


```{r}
rescale = function(x){
  rng = range(x, na.rm = T, finite = T)
  (x - rng[1])/ (rng[2] - rng[1])
}

x = c(1:10, Inf)

rescale(x)
```
Exercise
```{r}
#rewrite the function so that infinite is scaled to 1
rescale = function(x){
  rng = range(x, na.rm = T, finite = T)
  (x - rng[1])/ (rng[2] - rng[1])
  x[x == Inf] = 1
  x
}

x = c(1:10, NA, NA)
rescale(x)

#practice following snippets into functions
mean(is.na(x))

new_mean = function(x){
  y = x[is.na(x)]
  length(y)/length(x)
}
new_mean(x)
mean(is.na(x))

standardize = function(x){
  y = x[!is.na(x)]
  sum_y = sum(y)
  x/sum_y
}

sum(standardize(x), na.rm = T)

x = c(NA, 1, 2, 5, 'x', NA)
y = c(NA, 1,2,3,'y', 0)

is.na(x) & is.na(y)
```

if else exercise
```{r}
library(lubridate)



greet = function(time = now()){
  if (hour(time) >= 12){
    print('Good afternoon')
  }
  else if (hour(time) >= 6 & hour(time) <12){
    print('Good morning')
  }
  else {
    print('goodnight')
  }
}

greet(time = hms('08:15:02'))


fizzbuzz = function(num){
  if (num%%3 == 0 & num%%5 == 0){
    print('fizzbuzz')
  }
  else if (num%%3 == 0){
    print('fizz')
  }
  else if (num%%5 == 0){
    print('buzz')
  }
  else{
    num
  }
}

fizzbuzz(16)

#what happens if you use switch in numeric. 
x = c('apple','banana','cantaloupe')
switch(1, x)
switch(3, 'apple','banana','cantaloupe')

x = 'e'

switch(x,
       a = ,
       b = 'ab',
       c =,
       d = 'cd')


```


dot dot dot - many functions takes arbitrary inputs
```{r}
commas = function(...){
  str_c(..., collapse = ',')
}

commas(letters[1:25])

commas(letters[1:5], collapse = '-')
```

pipeable functions - there are two main types of pipeable functions

transformation functions - theres a clear primary object that is passed as the first argument, and modified version is returned by the function

side effect functions - called to perform an action like drawing a plot or saving a file. should invisible return the first argument so they are not printed by default, but are still pipeable

```{r}
show_missings = function(df){
  n = sum(is.na(df))
  cat('Missing values: ', n, '\n', sep = '')
  invisible(df)
}

show_missings(mtcars)

mtcars %>%
  show_missings() %>%
  mutate(mpg = ifelse(mpg<20, NA, mpg)) %>%
  show_missings()

?ifelse


```



















