---
title: "Iteration with purrr"
output: html_notebook
---


```{r}
library(tidyverse)
library(nycflights13)
```


```{r}
df = tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df

#get the median of each column
median(df$a)
median(df$b)
median(df$c)
median(df$d)

#we could use a for loop
output = vector('double',ncol(df))
output

for (i in seq_along(df)){
  output[[i]] = median(df[[i]])
}

output
```
Every for loop has three components

output vector - vector('double', length(x)). Before you start the loop, allocate sufficient space for output. This is very important for efficiency.

sequence - i in seq_along(df) - determines what to loop over. each run of for loop will assign i to a different value in seq_along(df). it is a safe version of 1:length(df). the difference is taht if you have a zero length in vector, seq along does the right thing

body - output[[i]] = median(df[[i]]) - this is the code that does the work. 


Exercise
```{r}
# compute the mean of every column in mtcars
?vector
otuput = vector('double', length = ncol(mtcars))

for (i in seq_along(mtcars)){
  output[[i]] = mean(mtcars[[i]])
}

output

#determine each ttype of column in flights
flights
output = vector('character', length = ncol(flights))

for (i in seq_along(flights)){
  output[[i]] = typeof(flights[[i]])
}

output

#compute the number of unique values in each column of iris
iris

output = vector('double', length = ncol(iris))

for (i in seq_along(iris)){
  output[[i]] = n_distinct(iris[[i]])
}

output

# generate 10 random normals for each mu = -10, 0, 10 and 100

mu = c(-10, 0, 10, 100)
output = vector('list', length(mu))

?rnorm
for (i in seq_along(mu)){
  output[[i]] = rnorm(10, mean = mu[[i]])
}


```

Exercise
```{r}
#eliminate the for loops by taking advantage of existing functions that work with vectors

out = ''
for (x in letters){
  out = str_c(out,x)
}

?str_c
str_c(out,letters, collapse = '')


#2
std_dev = 0
x = sample(100)
for (i in seq_along(x)){
  std_dev = std_dev + (x[i]-mean(x))^2
}
std_dev = sqrt(std_dev/ length(x)-1)
std_dev
sd(x)


```

For loop variations
- modifying an existing object instead of creating new object
- looping over name or values instead of indicides
- handling outputs of unknown lengths
- handling sequencing of unknown lengths

Modifying an existing object
```{r}
df = tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

rescale = function(x){
  rng = range(x, na.rm = T)
  (x - rng[1])/ (rng[2] - rng[1])
}

df$a = rescale(df$a)
df$b = rescale(df$b)

#using for loops
for (i in seq_along(df)){
  df[[i]] = rescale(df[[i]])
}
```


Looping patterns - aside from seq_along(xs) and extracting value with x[[i]] there are other two forms
loop over elements - for (x in xs)
loop over names - for (nm in names(xs)), this gives you a name to access value with x[[nm]]

Unknown output length - you might be tempted to solve this problem by progressively growing the vector
```{r}
means = c(0,1,2)

output = double()

for (i in seq_along(means)){
  n = sample(100,1)
  output = c(output, rnorm(n, means[[i]]))
}

str(output)

# a better solution is to save results in a list then combine to a single vector after the loop is done
out = vector('list', length(means))

for (i in seq_along(means)){
  n = sample(100,1)
  out[[i]] = rnorm(n, mean = means[[i]])
}
str(out)
str(unlist(out))
unlist(out)
```

Unknown sequence length - you can use while loop

while(condition){
body
}

for (i in seq_along(x)){
body
}

is equal to

i = 1
while(i <= length(x)){

i+1
}
```{r}
#use a while loop to get three heads in a row
flip = function(){
  sample(c('T','H'),1)
}

flips = 0
nheads = 0

while(nheads < 3){
  if(flip() == 'H'){
    nheads = nheads + 1
  }
  else{
    nheads = 0
  }
  flips = flips + 1
}
flips
```

Exercises
```{r}
files <- dir(getwd(), pattern = "\\.csv$", full.names = TRUE)
files

#pre allocate list
df_list <- vector("list", length(files))

for (i in seq_along(files)){
  df_list[[i]] = read_csv(files[[i]])
}

print(df_list)

#use bindrows to combine into single df
df = bind_rows(df_list)

#what happens if you for (nm in names(x)) and x has no names

x = c(11,12,13)
print(names(x))

for (nm in names(x)){
  print(nm)
  print(x[[nm]])
}

x <- c(a = 11, a = 12, c = 13)

```

For loops vs functionals - for loops are not as important in R as they are in other languages because R is a functional programming language. This means that it is possible to wrap up loops in a function and call that function instead of using loop directly


```{r}
df = tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

col_summary = function(df, fun){
  out = vector('double', length = length(df))
  for (i in seq_along(df)){
    out[[i]] = fun(df[[i]])
  }
  out
}

col_summary(df, mean)
```
The idea of passing a function to another function is extremely a powerful idea. 

Purrr package eliminates the need for for loops. The goal of purrr allow to break common list manipulation challenges to independent pieces

?apply()

Map functions

map() - makes a list
map_lgl() = makes a logical vector
map_int() - makes integer vector
map_double() - makes a double vector
map_chr() - makes a character vector

each function takes a vector as an input, applies function to each piece and returns a new vector of same length. 
```{r}

df
?map_dbl
map_dbl(df, mean)
map_dbl(df, mean)

#can also use with pipe
df %>%
  map_dbl(median)

#map functions also preserves names
x = list(x = 1:3, y = 4:5)
map_int(x, length)


```

Shortcuts - there are few shortcuts you can use with .f to save a little typing. Imagine you want to fit a linaer model to each group in the dataset. The following example splits up mtcars into three pieces for each cylinder and fits the same linear model to each piece

```{r}
mtcars %>%
  split(.$cyl) %>%
  map(function(df) lm(mpg~wt, data = df))

models = mtcars %>%
  split(.$cyl) %>%
  map(~lm(mpg ~ wt, data = .))

```

Used . as a pronoun it refers to the current list of element. when looking at many models, you might want to extract a summary statistics like R^2. To do that we need first to run summary() and extract component called r.squared. We could do that using the short hand for anonymous functions

```{r}
models %>%
  map(summary) %>%
  map_dbl(~.$r.squared)

models %>%
  map(summary) %>%
  map_dbl('r.squared')

# you can also use an integer to select elements by position

x = list(list(1,2,3), list(4,5,6), list(7,8,9))
x %>%
  map_dbl(2)

?map_dbl
```


Exercise
```{r}
#compute every mean of column in mtcars
mtcars %>%
  map(mean)

flights %>%
  map(typeof)

iris %>%
  map(n_distinct)

mu = c(-10,0,10,100)
mu %>%
  map(~rnorm(10, mean = .))

#check wether columns are factor

diamonds %>%
  map(is.factor)
map_lgl(diamonds, is.factor)

#eliminate anonymous function
x <- split(mtcars, mtcars$cyl)
map(x, function(df) lm(mpg ~ wt, data = df))

#using the ~shortcut
mtcars %>%
  split(.$cyl) %>%
  map(~lm(mpg~wt, data = .))

#or create a function
run_reg = function(df){
  lm(mpg~wt, data = df)
}
mtcars %>%
  split(.$cyl) %>%
  map(run_reg)

```

Dealing with failure
safely() - takes a function and returns a modified version. it will return a list of two elements: original result and error object
```{r}

safe_log = safely(log)

str(safe_log(10))

str(safe_log('a'))

x = list(1, 10, 'a')
y = x %>%
  map(safely(log))
str(y)

#transpose to see all errors and outputs
transpose(y)

#other variations are possibly and quietly
```

Mapping over multiple arguments
that is the job of map2 or pmap
```{r}
mu = list(5, 10, -3)
mu %>%
  map(rnorm, n = 5) %>%
  str()

#what if we want to add sd

sigma = list(1,5,10)
?rnorm
map2(mu, sigma, rnorm, n = 10)

#pmap

n = list(1,3,5)
args1 = list(n, mu, sigma)

args1 %>%
  pmap(rnorm) %>%
  str()

args2 = list(mean = mu, sd= sigma, n = n)
args2 %>%
  pmap(rnorm)

#can store to dataframe
params = tribble(
  ~mean, ~sd, ~n,
  5,1,1,
  10,5,3,
  -3,10,5
)

params %>%
  pmap(rnorm)
```

Invoking different functions
```{r}
f = c('runif', 'rnorm','rpois')
params = list(
  list(min = -1, max = 1),
  list(sd = 5),
  list(lambda = 10)
)

params
#use invoke map
invoke_map(f, params, n = 5)

#use tribble to match easier

sim = tribble(
  ~f, ~params,
  'runif', list(min = -1, max = 1),
  'rnorm', list(sd = 5),
  'rpois', list(lambda = 10)
)

sim %>%
  mutate(sim = invoke_map(f, params, n = 10)) %>%
  select(sim)


```

Other patterns of For loops

predicate functiosn return either True or False.
keep() and discard() keep elements of input where the predicate is T or F
```{r}
iris %>%
  keep(is.factor) %>%
  str()

iris %>%
  discard(is.factor)
```


Some() and every() determine if the predicate is true for any or for all elements
```{r}
x = list(1:5, letters, list(10))
x %>%
  some(is_character)

x %>%
  every(is_vector)
```

detect() finds the first elmement where predicate is true and detect index returns its position
```{r}
x = sample(10)

x %>%
  detect(~. >5)
```


Reduce and accumulate - 
```{r}
age = tibble(name = 'John', age = 30)
sex = tibble(name = c('Mary','John'), sex = c('M','F'))
trt = tibble(name = 'Mary', treatment = 'A')

dfs = list(age, sex, trt)

dfs %>% reduce(full_join)


#or maybe you have a list of vectors and you want to find the intersection

vs = list(
  c(1,3,5,6,10),
  c(1,2,4,5,6,10),
  c(1,2,3,4,9,10)
)

vs %>%
  reduce(union)


```



