---
title: "Factors with forcats"
output: html_notebook
---

Prerequisites

```{r}
library(tidyverse)
library(forcats)
```


Creating factors
```{r}
x1 = c('Dec','Apr','Jan','Mar')

x1 #not sorted

#can fix with levels
month_levels = c('Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec')
factor(x1, levels = month_levels)

#any values not in the set will be silently converted to NA
x2 = c('Jam','Feb','Mar','APr')
factor(x2, levels = month_levels)

#if you want an error you can use parse_factor
y2 = parse_factor(x2, levels = month_levels)

#sometimes youd prefer the order of the levels to match the order of appearance in the data. you can do that using unique()

factor(x1, levels = unique(x1))

```

Use general society survey to learn forcats
```{r}
gss_cat
str(gss_cat)

ggplot(data = gss_cat, mapping = aes(x = race))+
  geom_bar()

gss_cat %>%
  count(race)
#by default ggplot2 will drop levels that dont have anyvalue. You can force them to display
ggplot(gss_cat, aes(race))+
  geom_bar()+
  scale_x_discrete(drop = F)


```
Exercise
```{r}
#explore the distribtuion of rincome, what makes the default barchart hard to understand. How could you improve the plot

gss_cat %>%
  select(rincome) %>%
  count(rincome) %>%
  ggplot( aes(x = rincome, y = n))+
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  coord_flip()

#improvign the plots
gss_cat %>%
  filter(!rincome %in% c('Not applicable')) %>% # remove not applicable
  mutate(rincome = fct_recode(rincome, 'less than $1000' = 'Lt $1000')) %>%
  mutate(rincome_na = rincome %in% c('Refused',"Don't know",'No answer')) %>%
  ggplot(aes(x = rincome, fill = rincome_na))+
  geom_bar()+
  coord_flip()+
  scale_y_continuous('Number of Respondents', labels = scales::comma)+
  scale_x_discrete('Income')+
  scale_fill_manual(values = c("FALSE" = 'black', 'TRUE' = 'gray'))+
  theme(legend.position = 'None')

#most common relig and most common partyid
gss_cat %>%
  count(relig) %>%
  arrange(desc(relig))

gss_cat %>%
  count(partyid) %>%
  arrange(desc(n))

#which relig does denomination apply to. 
gss_cat %>%
  filter(!denom %in% c('Not applicable','Other',"Don't know",'No answer','No denomination')) %>%
  count(relig)

#scatterplot between demnom and relig
gss_cat %>%
  count(relig, denom) %>%
  ggplot(aes(relig, denom, size = n))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Modifying factor order - it is often useful to change the factor level in a visualization. Example, you want to explore the average hours spent in tv across religion
```{r}
relig = gss_cat %>%
  group_by(relig) %>%
  summarise(mean_hrs = mean(tvhours, na.rm = T))

gss_cat %>%
  group_by(relig) %>%
  summarise(mean_hrs = mean(tvhours, na.rm = T)) %>%
  ggplot(aes(mean_hrs, relig))+
  geom_point()


```
It is hard to interpret plot because there was no order. we can improve it by fct_reorder
takes 3 arguments:
f - factor you want to modify
x - a numeric vector you want to use to reorder levels
fun - function to use if there is multiple values of x. default is median

```{r}
gss_cat %>%
  group_by(relig) %>%
  summarise(mean_hrs = mean(tvhours, na.rm = T)) %>%
  ggplot(aes(mean_hrs, fct_reorder(relig, mean_hrs)))+
  geom_point()

#as you start making complicated transformations, move out of aes and use separate mutate 
relig %>%
  mutate(relig = fct_reorder(relig, mean_hrs)) %>%
  ggplot(aes(mean_hrs, relig))+
  geom_point()
```
create similar plot looking on how average age varies across reported income level
```{r}
rincome = gss_cat %>%
  group_by(rincome) %>%
  summarize(age = mean(age, na.rm = T),
            tvhours = mean(tvhours, na.rm = T),
            n = n())

ggplot(rincome, aes(age, fct_reorder(rincome, age)))+
  geom_point()
```

Here reordering is not a good idea. Reserve fct_roeorder for factors whose levels are not arbitrarily ordered


fct_relevel() takes a factor f and then any number of levels that you want to move on the front
```{r}
ggplot(rincome, aes(age, fct_relevel(rincome, 'Not applicable', 'Refused')))+
  geom_point()
```

Another type of reordering is useful when coloring lines on plot. fct_reorder2 reorders the factor by the y values associated with the largest x values.
```{r}
?fct_reorder2
by_age = gss_cat %>%
  filter(!is.na(age)) %>%
  group_by(age, marital) %>%
  summarise(total = n()) %>%
  mutate(prop = total/ sum(total))

by_age
ggplot(by_age, aes(age, prop, color = marital))+
  geom_line(na.rm = T)

ggplot(by_age, aes(age, prop, color = fct_reorder2(marital,age,prop)))+
  geom_line() +
  labs(color = 'marital')
```

For barplots you can use fct_infreq to order levels in increasing frequency
```{r}
gss_cat %>%
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(marital))+
  geom_bar()
```

Exercises
```{r}
#there are suspicously high number of tv hrs. is the mean a good summary
gss_cat %>%
  filter(!is.na(tvhours)) %>%
  ggplot(aes(tvhours))+
  geom_histogram(binwidth = 1)

#for each factor in gss_cat identify whether the order of the levels are arbitrary or princples
str(gss_cat)

levels(gss_cat$marital)

levels(gss_cat$relig)
gss_cat %>%
  ggplot(aes(relig))+
  geom_bar()
```

Modifying factor levels - allows you to collapse levels for high level displays
```{r}
gss_cat %>%
  count(partyid)

#these levels are strong and incosistent. lets tweak
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
                              'Republican, strong' = 'Strong republican',
                              'Republican, weak' = 'Not str republican',
                              'Independent, near rep' = 'Ind,near rep',
                              'Independent, near dem' = 'Ind,near dem',
                              'Democrat, weak' = 'Not str democrat',
                              'Democrat, strong' = 'Strong democrat'
  )) %>%
  count(partyid)

#can assign multiple old levels to the same new level
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
                              'Republican, strong' = 'Strong republican',
                              'Republican, weak' = 'Not str republican',
                              'Independent, near rep' = 'Ind,near rep',
                              'Independent, near dem' = 'Ind,near dem',
                              'Democrat, weak' = 'Not str democrat',
                              'Democrat, strong' = 'Strong democrat',
                              'Other' = 'No answer',
                              'Other' = "Don't know",
                              'Other' = 'Other party'
  )) %>%
  count(partyid)

#if you want to collapse a lot of levels you can use fct_collapse which is a variant of fct_recode

gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
                                other = c('No answer',"Don't know",'Other party'),
                                rep = c('Strong republican','Not str republican'),
                                ind  = c('Ind,near rep','Independent','Ind, near dem'),
                                dem = c('Not str democrat','Strong democrat'))) %>%
  count(partyid)

#sometimes you want to lump small group to make plot or table easier
gss_cat %>%
  mutate(relig = fct_lump(relig)) %>%
  count(relig)

#the default behavior is to lump together the smallest group. we can use n parameter to specify how many groups we want

gss_cat %>%
  mutate(relig = fct_lump(relig, n = 6)) %>%
  count(relig, sort = T) %>%
  print(n = Inf)
```


Exercise
```{r}
#how many proportions of people identifying democrat, republican, or independent changed overtime
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
                                other = c('No answer',"Don't know",'Other party'),
                                rep = c('Strong republican','Not str republican'),
                                ind  = c('Ind,near rep','Independent','Ind, near dem'),
                                dem = c('Not str democrat','Strong democrat'))) %>%
  count(year, partyid) %>%
  group_by(year) %>%
  mutate(p = n / sum(n)) %>%
  ggplot(aes(x = year, y = p, color = fct_reorder2(partyid,year,p)))+
  geom_line()+
  geom_point()+
  labs(color = 'party id')

#how could you collapse rincome to small set of categories
gss_cat %>%
  count(rincome)
gss_cat %>%
  mutate(rincome = fct_collapse(rincome,
                                Other = c('Not applicable','Refused','No answer',"Don't know"))) %>%
  ggplot(aes(rincome))+
  geom_bar()+
  coord_flip()
```


```{r}
library(stringr)
gss_cat %>%
  mutate(
    rincome =
      fct_collapse(
        rincome,
        `Unknown` = c("No answer", "Don't know", "Refused", "Not applicable"),
        `Lt $5000` = c("Lt $1000", str_c(
          "$", c("1000", "3000", "4000"),
          " to ", c("2999", "3999", "4999")
        )),
        `$5000 to 10000` = str_c(
          "$", c("5000", "6000", "7000", "8000"),
          " to ", c("5999", "6999", "7999", "9999")
        )
      )
  ) %>%

```



















