---
title: "Dates and time with lubridate"
output: html_notebook
---


```{r}
library(tidyverse)
library(lubridate)
library(nycflights13)
```

Creating date/times
there are three types of date time data. 
A date - <date>
time within the day - <time>
date-time - an instant in time <POSIXct>
```{r}
today()

now()
```
From strings

```{r}
#aside from parsing, another approach is to use lubridate
ymd('2017/01-31')

mdy('Jan 31, 2017')

dmy('31/Jan-2017')

ymd(20170131)

#to create a date time add h m or s

ymd_hms('2017-01-31 20:11:59')
mdy_hm('01/31/2017 08:01')
```
From individual components
```{r}
flights %>%
  select(year, month, day, hour, minute)

#use make_date for dates
flights %>%
  select(year, month, day, hour, minute) %>%
  mutate(
    departure = make_datetime(year,month,day,hour,minute)
  )

#pull out hour and minute at time

make_date_time100 = function(year,month,day,time){
  make_datetime(year,month,day,time%/%100, time%% 100)
}

flights_dt = flights %>%
  filter(!is.na(dep_time), !is.na(arr_time)) %>%
  mutate(
    dep_time = make_date_time100(year,month,day,dep_time),
    arr_time = make_date_time100(year,month,day,arr_time),
    sched_dep_time = make_date_time100(year,month,day,sched_dep_time),
    sched_arr_time = make_date_time100(year,month, day, sched_arr_time)
    ) %>%
  select(origin, dest, ends_with('delay'), ends_with('time'))

flights_dt

#visualize the distribution of departure time across the year
flights_dt %>%
  ggplot(aes(dep_time))+
  geom_freqpoly(binwidth = 86400) #daily

#within a singleday
flights_dt %>%
  filter(dep_time < ymd(20130102)) %>%
  ggplot(aes(dep_time))+
  geom_freqpoly(binwidth = 600) #every 10 mins
```
From other types - you may want to switch between date time and date. that is the job of as_date_time and as_date()

```{r}
as_datetime(today())
as_date(now())
```
Exercises
```{r}
#What happens if you parse a string that contains invalid dates

ymd(c('2010-1010','banaa'))

?today()

#use lubridate functions to parse the ff
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14"

mdy(d1)
ymd(d2)
dmy(d3)
mdy(d4)
mdy(d5)
```

Date time components - you can pull out parts of date using year(),month(),mday(), yday(), wday(), hour(), minute(), sec()
```{r}
date_time = ymd_hms('2016-07-08 12:34:56')
year(date_time)
month(date_time)
mday(date_time)
yday(date_time)
wday(date_time)

#set label = T to return abbreviated name
month(date_time, label = T)
wday(date_time, label =  T)

#use wday() to see flights during the weekdays

flights_dt %>%
  mutate(wday = wday(dep_time, label = T)) %>%
  ggplot(aes(wday))+
  geom_bar()

#theres an interesting pattern if we look at average deparature delay by minute within the hour. it looks like the flights leaving in minutes 2030 have lower delays

flights_dt %>%
  mutate(minutes = minute(dep_time)) %>%
  group_by(minutes) %>%
  summarize(avg_delay = mean(arr_delay, na.rm = T), 
            n = n()) %>%
  ggplot(aes(minutes, avg_delay))+
  geom_line()

#if we look at scheduled departure time we dont see such a strong pattern
flights_dt %>%
  mutate(minute  = minutes(sched_dep_time)) %>%
  group_by(minute) %>%
  summarize(
    avg_delay = mean(arr_delay, na.rm = T),
    n = n()
  ) %>%
  ggplot(aes(minute, avg_delay))+
  geom_line()
```

Rounding - an alternative approach is rounding date to nearby unit of time with floor_date(), round_date(), ceiling_date(). 
```{r}
flights_dt %>%
  count(week = floor_date(dep_time, 'week')) %>%
  ggplot(aes(week, n))+
  geom_line()

```

Setting components - you can use each accessor function to set components of date/time
```{r}
datetime = ymd_hms('2016-07-08 12:34:56')
datetime

year(datetime) = 2020
datetime

month(datetime) = 'Jan'
datetime

hour(datetime) = hour(datetime) + 3
datetime

#you can use update
update(datetime, year = 2020, month =2, day = 2, hour = 2)

#you can use update() to show distribution of flights across the course of day for everyday of the year
?update
flights_dt %>%
  mutate(dep_hour = update(dep_time, yday = 1)) %>%
  ggplot(aes(dep_hour))+
  geom_freqpoly(binwidth= 300)
```
Exercise
```{r}
#how does distribution of flight times within a day change over the course of the year. #normalize because february has fewer days
flights_dt %>%
  filter(!is.na(dep_time)) %>%
  mutate(dep_hour = update(dep_time, yday = 1)) %>%
  mutate(month = factor(month(dep_time))) %>%
  ggplot(aes(dep_hour, color = month))+
  geom_freqpoly(aes(y = ..density..),binwidth = 3600)

#compare dep_time, sched_dep_time, dep_delay. are they consistent?

flights_dt %>%
  mutate(dep_time_ = sched_dep_time + dep_delay * 60) %>%
  filter(dep_time != dep_time_ )

#on what week of day should you leave to minimize chance of delay or lowest average departure and delay

flights_dt %>%
  mutate(dow = wday(sched_dep_time)) %>%
  group_by(dow) %>%
  summarize(
    dep_delay = mean(dep_delay, na.rm = T),
    arr_delay = mean(arr_delay, na.rm = T)
  ) %>%
  gather(dep_delay, arr_delay, key = 'delay', value = 'time') %>%
  ggplot(aes(dow, time, fill = delay))+
  geom_col(position = 'dodge')


flights_dt %>%
  mutate(wday = wday(sched_dep_time, label = T)) %>%
  group_by(wday) %>%
  summarize(ave_dep_delay = mean(dep_delay, na.rm = T)) %>%
  ggplot(aes(wday, ave_dep_delay))+
  geom_col()+
  labs(y = 'Average dep delay', x = 'weekday')
``` 

Time spans - three important classes to represent time spans
durations - exact number of seconds
periods - like weeks or months
intervals - represent a start and ending point
```{r}
#if you subtract two dates, you get a time diff
age = today() - ymd(19951205)
as.duration(age) #convert to duration

#constructors of duration
dseconds(15)
dminutes(600)
dhours(c(12,24))
ddays(0:5)
dweeks(3)

#can add durations
2 * dyears(1)

tomorrow = today() + ddays(1)
tomorrow

#periods - are time spans in human time

one_pm = ymd_hms('2016-03-12 13:00:00')
one_pm + days(1)

#constructor of periods
minutes(15)
hours(c(12,24))
months(1:6)


#use period to fix flight dates. some planes appear to have arrived at their destination before they departed
flights_dt %>%
  filter(arr_time < dep_time)

#these are overnight flights. can fix by adding 1 day to arrival of each overnight flight
flights_dt = flights_dt %>%
  mutate(
    overnight = arr_time < dep_time,
    arr_time = arr_time + days(overnight * 1),
    sched_arr_time = sched_arr_time + days(overnight * 1)
  )

flights_dt %>%
  filter(arr_time < dep_time)
```

Intervals
```{r}
years(1)/days(1)

#if you want more accurate measurement youll have to use interval
next_year = today() + years(1)
next_year

(today() %--% next_year)/ ddays(1)
```

Exercises
```{r}
#why is there months() but no dmonths()
#because of differences in moths

flights_dt %>%
  select(overnight) %>%
  mutate(sum = days(overnight * 1))

#create a vector of dates giving the first day of everymonth in 2015

year = years(2015)
year + months(c(1,2,3,4,5,6,7,8,9,10,11,12)) + days(c(1))

#create a vector of dates giving the first day of everymonth this year
floor_date(today(), unit = 'year') + months(0:11)
```



















