---
title: "Data Visualization with GGplot"
output: html_notebook
---

#installing packages
install.packages('tidyverse') #packages for data 
install.packages(c('nycflights13','gapminder','Lahman')) #data

Prerequisites
```{r}
library(tidyverse)
```

```{r}
mpg # the dataset

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

```
The plot shows negative relatiosnhip between engine size (displ) and fuel efficiency (hwy). 

template
ggplot(data = <DATA>) +
  <GEOM FUNCTION>(mapping = aes (<MAPPINGS>))

#exercise
```{r}
#run ggplot 
ggplot(data = mpg) # shows a blank frame

#rows and columns of mtcards
dim(mtcars) 

#what does variable drv describe
?mtcars

#make a scatterplot of hwy vs cyl
ggplot(data = mpg) +
  geom_point(mapping = aes(y = hwy, x = cyl))

```
the plot is not useful because it is not quantitative

#Aesthetic Mapping
```{r}
#adding color like hue
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, color = class))

#adding size
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, size = class)) #size depends on class


#can use alpha instead for class
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

#or shape
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))

#set the arguments of geom_points manually. set it outside aes
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy), color = 'blue', size= 3, shape = 22, fill = 19, pch = 2)
```

#Exercise
```{r}
#which variables in mpg are categorical
summary(mpg)
map_lgl(mpg, is.factor)

#how do aesthetics behave differently to categorical vs continous
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, color = cyl))

#map same variable to multiple aesthetics
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, color = cyl, alpha = cyl))

#what does stroke aesthetic doe
?geom_point
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, stroke = cyl), color = 'red')

#what happens if u map other than variable name (color = displ < 5)
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, color = displ<5))
#True or False value
```
#Facets
```{r}
#use facet_wrap(). The variable that you pass to facet_wrap should be discrete
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))+
  facet_wrap(~class, nrow = 2)

#facet_grid() for two variables

ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))+
  facet_grid(class~drv)
```
Exercise
```{r}
summary(mpg)
#facet on continous variable will produce an error
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))+
  facet_wrap(~manufacturer)

#what do the empty cells mean - means no value
ggplot(data = mpg)+
  geom_point(mapping = aes(x = drv, y = cyl))

ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))+
  facet_grid(drv~cyl)

#what does . do in facetgrid
#facet grid into rows yung drive
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))+
  facet_grid(drv ~ .) 

#facetgrid into columns yung clas
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))+
  facet_grid(. ~ cyl)

#what are the advantages of choosing facet over color
#for larger dataset better to choose color
#facet wrap or grid is better if u want to see individually and check the differences
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))+
  facet_wrap(~class, nrow = 2)

#what does nrow and ncolumn do
?facet_wrap

```

Geometric objects
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg)+
  geom_smooth(mapping = aes(x = displ, y = hwy)) #smoothed line

#not every aesthetic works with geom

ggplot(data = mpg)+
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))

#use group to draw multiple variables. group does not add legend

par(mfrow = c(3,1))
ggplot(data = mpg)+
  geom_smooth(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

ggplot(data = mpg)+
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv), show.legend = F)

#display multiple geoms in the same plot
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))+
  geom_smooth(mapping = aes(x = displ, y = hwy))

#to avoid many typings

ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point()+
  geom_smooth()

#if you place mapping inside geomfunction, it will treat as local mappings for that layer.
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(color = class))+
  geom_smooth()

#can also specify data for each layer. here, the smooth line displays subset of mpg. The geom_smooth data overrides data global argument on that layer only

ggplot(data = mpg, mapping = aes(x = displ, y  = hwy))+
  geom_point(mapping = aes(color = class))+
  geom_smooth(data = filter(mpg, class == 'subcompact'), se = FALSE)

```

Exercise
```{r}
#what geom would draw a linechart, boxplot, a histogram, area chart
?mpg

ggplot(data = mpg)+
  geom_line(mapping = aes(x = year, y =hwy, color = class))

ggplot(data = mpg)+
  geom_boxplot(mapping = aes(x = drv, y = hwy))

ggplot(data = mpg)+
  geom_histogram(mapping = aes(x = cty))

ggplot(data = mpg)+
  geom_area(mapping = aes(x = cty, y = hwy))

#predict the output of code

ggplot(data = mpg, mapping = aes(x= displ, y = hwy, color = drv))+
  geom_point()+
  geom_smooth(se = F) #se stands for error

#what does show.legend do

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv))+
  geom_point(show.legend = F)

#check if there is difference between the two
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point()+
  geom_smooth()

ggplot()+
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))

#recreate the R codes necessary to generate the graphs
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(size = 5))+
  geom_smooth(se = F, mapping = aes(group = drv))

ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(color = drv))+
  geom_smooth(se = F, mapping = aes(linetype = drv), size = 3)

```

Statistical transformations
```{r}
?diamonds
head(diamonds)

#barplot
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut))

#recreate barplot using stat.count
ggplot(data = diamonds)+
  stat_count(mapping = aes(x = cut))

####Three reasons you might want to use stat explicitly

#the height is already presented in the data
demo = tribble(
  ~a, ~b,
  'bar1', 20,
  'bar2',30,
  'bar3',40
)

ggplot(data = demo)+
  geom_bar(mapping = aes(x= a, y = b), stat = 'identity')

#override default mapping transformed variables to aesthetics, display proportion rather than count
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))

#you might want to draw attention to statistical transformation in your code, You might want to use stat_summary()
ggplot(data = diamonds)+
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )

?stat_summary
```

#Exercise
```{r}
#what is the default geom associated with stat summary. how could you rewrite the previous plot to use geomfucntion instead of stat function
?geom_line
ggplot(data = diamonds)+
  geom_line(mapping = aes(x = cut, y = depth))

#what does geom col do, how is it different from geombar
?geom_col

ggplot(data = demo)+
  geom_col(mapping = aes(x = a, y = b))

#what does stat smooth do, what parameters control its behavior
?stat_smooth
ggplot(data = mpg)+
  stat_smooth( mapping = aes(x = displ, y = hwy))

#in proportion barchart we set the group to 1. Why?
#what are wrong in these graphs
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))

ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, fill = color))
```

Position adjustments
```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, color = cut))

#add color using fill
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, fill = cut))

#stacked bar if the fill is another variable
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut, fill = clarity))

#if you dont want stacked bar you can use three other position options 'identity', 'dodge','fill'

#identity = will place each object exactly where it falls in the context of the graph. not very useful for graphs because it overlaps them.

ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity))+
  geom_bar(alpha = .2,position = 'identity')

#fill = works like stacking but same height
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, fill = clarity), position = 'fill', alpha = .5)

#position = dodge places the overlapping objects side by side
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, fill =  clarity), position = 'dodge')

#add jitter to scatterplot to produce all points
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy), position = 'jitter')
```
Exercise
```{r}
#whats the problem with this plot, how can you improve it
jitter = ggplot(data = mpg, mapping = aes(x = cty, y = hwy))+
  geom_point(position = 'jitter')
jitter
#what parameter control the amount jittering
?geom_jitter
ggplot(data = mpg, mapping = aes(x = cty, y = hwy))+
  geom_jitter(height = 2)

#compare and contrast geom_jitter() with geom_count()
ggplot(data = mpg, mapping = aes(x = cty, y = hwy))+
  geom_count()

ggplot(mpg, mapping = aes(cty, hwy, color = drv))+
  geom_count(position = 'dodge')

#what is the default position for geom_boxplot

?geom_boxplot
ggplot(data = mpg, mapping = aes(x = drv, y = hwy))+
  geom_boxplot(position = 'identity')
```
coordinate systems
```{r}
#coord flip switches x and y axis. useful when there are long labels

ggplot(data = mpg, mapping = aes(x = class, y = hwy))+
  geom_boxplot()+
  coord_flip()

#coord polar uses polar coordinates
#the chart is in the object
bar = ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, fill = cut), show.legend = F, width = 1)+
  theme(aspect.ratio = 1)+
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()

```
#Exercise
```{r}
#turn a stacked barplot to piechart using coord polar

bar = ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, fill = clarity), position = 'fill')
bar+ coord_polar()

ggplot(mpg, aes(x = 1, fill = class))+
  geom_bar(position = 'fill')+
  coord_polar(theta = 'y')+
  scale_x_continuous(labels = NULL)

#if you want to make multiple levels
ggplot(mpg, aes(x = as.factor(cyl), fill = class))+
  geom_bar(position = 'fill')+
  coord_polar(theta = 'y')

#what does labs() do - for labels
?labs

#what does the plot tell you about the relationship between city and highway mpg. why is coord fixed important. what does abline do
#abline does add straight line
?coord_fixed #fix aspect ratio

ggplot(data = mpg, mapping = aes(x = cty, y = hwy))+
  geom_point()+
  geom_abline()+
  coord_fixed()
  
```
#layered grammar of graphics

The format

ggplot(data = <DATA>)+
  <geom function>(
  mapping = aes(<mappings>),
  stat = <stat>,
  position = <position>
  )+
  <coordinate>+
  <facet function>


```{r}
my_variable = 10
my_variable

#tweak the following command so they can work

library(tidyverse)
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy))

filter(mpg, cyl == 8)
filter(diamonds, carat > 3)

63-29
103-69
```











