---
title: "Pipes with magrittr"
output: html_notebook
---

Piping alternatives - the point of pipe is to help you write code in a way thats easier. explore number of ways writing the same code. tell a story a bout a bunny named foo foo

- Intermediate step
- overwrite the original object many times
- compose function
- use the pipe
```{r}
library(magrittr)
library(tidyverse)
install.packages('pryr')

little_bunny()
#intermediate step
hop(foo_foo)

diamonds2 = diamonds %>%
  mutate(price_per_carat = price/carat)

pryr::object_size(diamonds)
pryr::object_size(diamonds2)

pryr::object_size(diamonds, diamonds2)

#the way pipe was written, the pipe wont work for two classes function
assign('x',10)

'x' %>% assign(100)
x # same value

#if you want to assign using pipe
env = environment()
'x' %>% assign(100, envir = env)
x
```
The other functions with this problem include get() and load()

WHEN NOT TO USE PIPES
- Your pipes are longer than 10 steps. In that case create objects with meaningful names. that will make debugging easier. 
- You have multiple inputs or outputs. If there is not one primary object being transformed but two or more objects being combined together, dont use the pipe
- you are starting to think about a directed graph with complex dependency. 

Other tools from magrittr
tee-pipe for a function to return an object
```{r}

rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()

rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
```

If you are working with functions that dont have dataframe-based API. You might find %$% useful. it explodes out variables in dataframe so that you can refer to them explicitly. This is useful with many functions in base R

```{r}
mtcars %$%
  cor(disp, mpg)
```

%<>% operator for assignment

```{r}
mtcars
mtcars = mtcars %>%
  transform(cyl = cyl*2)

?transform

mtcars %<>%
  transform(cyl = cyl *2)
```




























